Phonon polaritonics in two-dimensional materials

### 2019-01-13 paper.tex
final edits -- ready

% --- ABSTRACT ---
\begin{abstract}
    Extreme confinement of electromagnetic energy by phonon polaritons holds the promise of strong and new forms of control over the dynamics of matter. 
    To bring such control to the atomic-scale limit, it is important to consider phonon polaritons in two dimensional systems. Recent studies have pointed out that in two-dimensional systems, splitting between longitudinal and transverse optical (LO and TO) phonons is absent at the $\Gamma$ point.  
    Does this lack of LO--TO splitting imply the absence of a phonon polariton in polar monolayers? 
    Here, we derive a first-principles expression for the spatially nonlocal conductivity of a polar monolayer specified by the wavevector-dependent LO and TO phonon dispersions. In the long-wavelength (local) limit, we find a universal form for the conductivity in terms of the LO phonon frequency at the $\Gamma$ point,
    its lifetime, and the group velocity of the LO phonon. Our analysis reveals that the phonon polariton of the bulk is essentially replaced by the LO phonon of the two-dimensional system. For the specific example of hexagonal boron nitride, we estimate the confinement and propagation losses of the LO phonons, finding that high confinement and reasonable propagation quality factors coincide in regions which are difficult to detect with current near-field optical microscopy techniques. 
    Finally, we study the interaction of external emitters with two-dimensional hBN nanostructures, 
    finding extreme enhancement of spontaneous emission due to coupling with localized two-dimensional (2D) LO phonons, and the possibility of multi-mode strong and ultra-strong coupling between an external emitter and hBN phonons. This may lead to the design of new hybrid states of electrons and phonons based on strong coupling.

@@

% --- INTRODUCTION ---
Phonon-polaritons, hybrid quasiparticles of photons and optical phonons supported in polar materials, hold tremendous promise for deeply sub-diffraction control of electromagnetic fields at mid-infrared and terahertz frequencies. Qualitatively, phonon polaritons share many features with plasmon polaritons in conductors. Recently, it has been shown that phonon polaritons enable confinement of light to volumes  $\sim10^6$ times smaller than that of a diffraction-limited photon in free-space~\cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra,li2018infrared,ma2018plane}. Due to this remarkable confinement and their relatively high lifetimes---around picoseconds---phonon polaritons open new opportunities for vibrational spectroscopy~\cite{autore2018boron}, radiative heat transfer~\cite{hillenbrand2002phonon}, and control of dynamics in quantum emitters~\cite{kumar2015tunable,rivera2017making,kurman2018control,flick2018strong}. 
Thus far, extreme confinement of phonon polaritons has been achieved by the use of thin-films (or nanostructuring), which shrink the in- and out-of-plane wavelength of polaritons monotonically with decreasing film thickness~\cite{dai2014tunable,dubrovkin2018ultra}. A monolayer is the ultimate limit of this effect. Concerning the optical response, the transition from three-dimensional (3D) to two-dimensional (2D) polar materials is nontrivial, however, since in a polar monolayer, the LO--TO splitting that gives rise to phonon polaritons in 3D is absent at the $\Gamma$ point~\cite{sanchez2002vibrational,mele2002electric,serrano2007vibrational,sohier2017breakdown}. This raises a fundamental question about the nature of  electromagnetic modes in polar monolayers.

@@

\section{Optical response of optical phonons in two-dimensions}
In this section, we develop a theory of electromagnetic response due to optical phonons in 2D systems. The key response function of interest is the conductivity of the monolayer. To that end, we consider the response of the ions of the monolayer due to an electric potential $\phi$. For that case, the interaction Hamiltonian is
\begin{equation}
    H_{\mathrm{int}} = \int \dd^2x ~\rho \phi = -\int \dd^2x~ (\nabla\cdot\mathbf{P})\phi,
    \label{eq:coupling}
\end{equation} 
with $\rho$ the induced charge density and $\mathbf{P}$ the polarization density associated with the ionic motion. 
Within linear response theory, the induced polarization density can be straightforwardly evaluated from the displacement amplitude $\mathbf{u}_{\kappa}$ of every atom $\kappa$ within the unit cell. Specifically, to first order, the induced polarization is 
\begin{equation}
\mathbf{P} - \mathbf{P}^0 = \sum_\kappa(\boldsymbol{\nabla}_{\mathbf{u}_\kappa}\mathbf{P})\mathbf{u}_\kappa \equiv \frac{1}{\Omega}\sum_\kappa\mathbf{Z}_{\kappa}\mathbf{u}_{\kappa},
\end{equation} 
where $\mathbf{Z}_\kappa \equiv \Omega\boldsymbol{\nabla}_{\mathbf{u}_\kappa}\mathbf{P}$ is the Born effective charge tensor of ion $\kappa$ and $\Omega$ is the unit cell area. $\mathbf{P}^0$ is the equilibrium polarization in the absence of displacements, which is zero here. With this relation between polarization and ionic displacements, the interaction Hamiltonian in Equation~\eqref{eq:coupling} couples the scalar potential and the ionic displacements.  We consider the response of the monolayer to a potential of the form $\phi(\mathbf{r}) = \phi(\mathbf{q},\omega)\e^{\iu\mathbf{q}\cdot\mathbf{r}-\iu\omega t}$, where $\mathbf{q}$ is a 2D wavevector in the plane of the monolayer. Such a potential corresponds to a longitudinal electric field  $\mathbf{E}(\mathbf{r}) = \iu\mathbf{q}\phi(\mathbf{r})$.

In what follows, we assume the validity of the random-phase approximation (RPA) in calculating Coulombic interactions between ions in the polar lattice. Within the RPA, these Coulombic interactions are accounted for by taking the induced polarization $\mathbf{P}(\mathbf{q},\omega)$ to be proportional \emph{total} electric field, $\mathbf{E}_{\mathrm{tot}}(\mathbf{q},\omega)$, defined to be the sum of the externally applied electric field and the electric field created by the induced polarization. The polarization and total-field are connected by the polarization-polarization response (tensor) function  $\boldsymbol{\Pi}(\mathbf{q},\omega)$ via
\begin{equation}
\mathbf{P}(\mathbf{q},\omega)  = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}_{\mathrm{tot}}(\mathbf{q},\omega),
\end{equation}
where boldfaced quantities refer to vectors or tensors as appropriate. The polarization-polarization response function is related to the conductivity via the relation $\boldsymbol{\sigma}(\mathbf{q},\omega) = -\iu\omega\epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)$. From the Kubo formula, it follows that the conductivity is
\begin{equation}\label{eq:2dsusceptibility}
    \boldsymbol{\sigma}(\mathbf{q},\omega) =  \frac{-\iu \omega}{ \Omega\mathcal{Z}}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+\iu 0^+}\Big(\e^{-\beta E_m}-\e^{-\beta E_n} \Big),
\end{equation}
where $m,n$ are states of the phononic Fock space of the monolayer, $\mathbf{P}_{mn}(\mathbf{q}) \equiv \sum_\kappa \mathbf{Z}_\kappa\langle m|\mathbf{u}_\kappa(\mathbf{q})|n\rangle$ are matrix elements of the polarization associated with phonon modes (where $\mathbf{u}_\kappa(\mathbf{q})$ is the Fourier transform of the phonon displacement operator), $E_{m}$ ($E_n$) is the energy of state $m$ ($n$), $\beta \equiv 1/k_{\text{\textsc{B}}}T$ is the inverse temperature, and $\mathcal{Z}$ is the grand partition function. We now evaluate the contribution of optical phonons to the polarization-polarization response in the low-temperature limit $T \ll \hbar\omega_{\mathrm{ph}}/k_{\mathrm{B}}$ with $\omega_{\mathrm{ph}}$ a characteristic optical phonon frequency.  Considering the long-wavelength (small wavevector) limit, and taking a material with long-wavelength isotropy, such as hBN, we only have to consider the $qq$-component in the response tensor, where $qq$ denotes a pair of directions parallel to the wavevector. Denoting $\sigma_{qq}$ as simply $\sigma$, we find that the conductivity is given by \cite{rivera2018ab}: 
\begin{equation}
    \sigma(\mathbf{q},\omega) = \frac{-\iu\omega}{\hbar\Omega} \frac{2\omega_{\mathbf{q},\mathrm{L}}}{\omega^2_{\mathbf{q},\mathrm{L}}-\omega^2-\iu\omega\tau^{-1}}|\hat{\mathbf{q}}\cdot\langle 1_{\mathbf{q},\mathrm{L}}|\mathbf{P}(\mathbf{q})|0_{\mathbf{q},L}\rangle|^2,
    \label{eq:polpolresponse}
\end{equation}
where L-subscripts denote longitudinal polarization, $|0_{\mathbf{q},\mathrm{L}}\rangle$ ($|1_{\mathbf{q},\mathrm{L}}\rangle$) denotes a state with no (one) longitudinal phonon of wavevector $\mathbf{q}$, and $\hat{\mathbf{q}}$ denotes a unit vector in the direction of $\mathbf{q}$. We have also phenomenologically included the phonon dissipation rate $\tau^{-1}$, consistently with  a relaxation-time prescription.  The frequency $\omega_{\mathbf{q},\mathrm{L}}$ in the denominator, as in the case of bulk phonons, is the frequency of the longitudinal phonon of wavevector $\mathbf{q}$ prior to considering LO--TO splitting \cite{BornHuang:1954}.
This is consistent with the fact that LO--TO splitting is a collective effect arising from Coulomb interactions and the fact that the equation above represents a single-particle susceptibility. Coulomb interactions are accounted for in the random phase approximation, and to include them in the single-particle response amounts to an uncontrolled double-counting.

@@

While the conductivity is the main electromagnetic quantity of interest for electrodynamics applications, we briefly state the form of the (2D) permittivity, as its zeros immediately yield the longitudinal modes of the system, which are the LO phonons. The permittivity within the RPA, denoted $\epsilon_{\textrm{RPA}}$ is related to the polarization-polarization response function via \cite{jablan2009plasmonics} $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{1}{2}q\Pi(q,\omega)$ and the conductivity via $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{\iu q\sigma(q,\omega)}{2\omega\epsilon_0}$. Here, $\epsilon_{\mathrm{env}}$ is the average permittivity of the bulk above and below the monolayer, and is added to take into account the polarization arising from these bulk materials. Note that we have neglected any intrinsic high-frequency screening in the monolayer itself, which is only relevant for wavevectors comparable to the inverse layer spacing between monolayers. When considering non-local corrections to the conductivity at these large wavevectors, these must be taken into account \cite{sohier2016two,thygesen2017calculating,sohier2017breakdown}. Based on Equation~\eqref{eq:conductivity_simplified}, the zeros $\omega_{\mathbf{q}}$ of the RPA dielectric function satisfy:
\begin{equation}
    \omega^2_{\mathbf{q}} - \omega^2_{\mathrm{TO}} = V(q)\frac{1}{\Omega}q^2\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\tilde{\mathbf{Z}}_{\kappa}\boldsymbol{\eta}_{\kappa}\Big|^2,
    \label{eq:zerosofresponse}
\end{equation}
  where $e\tilde{\mathbf{Z}}_\kappa \equiv \mathbf{Z}_\kappa$ and $V(q)$ is the Coulomb interaction in Fourier space, which in two dimensions, is given by $V(q) = \frac{e^2}{2\epsilon_0\epsilon_{\mathrm{env}}q}$. Given that the zeros of the dielectric function are associated with longitudinal modes, one expects that $\omega_{\mathbf{q}}$ is in fact the frequency of the LO phonon mode. This is consistent with the result of Ref.~\citenum{sohier2017breakdown}, in which \relax{{}\ }\citeauthor{sohier2017breakdown} show that in 2D polar materials, the extra restoring forces on LO phonons relative to TO phonons, due to the Coulomb interaction, lead to a wavevector-dependent LO--TO splitting and zero LO--TO splitting at the $\Gamma$ point of the Brillouin zone.

Given these results, we now re-express the conductivity explicitly in terms of the 2D phonon dispersion, and derive a universal form for the conductivity in the local ($q\rightarrow 0$) limit specified in terms of three parameters: the TO phonon frequency, the group velocity of the LO phonon at the $\Gamma$ point, and the damping rate. From Equation~\eqref{eq:zerosofresponse}, we can immediately write the conductivity as
  \begin{equation}
  \sigma(\mathbf{q},\omega) = \frac{-2\iu\epsilon_0\epsilon_{\mathrm{env}}\omega}{q}\frac{\omega^2_{\mathbf{q},\mathrm{LO}}-\omega^2_{\mathrm{TO}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
  \end{equation}
 In this expression, $\omega_L$, the LO phonon frequency prior to LO-TO splitting, has been re-named as $\omega_{\mathrm{TO}}$ , the transverse optical phonon frequency, because in the absence of LO--TO splitting, they are degenerate. This is done in order to make the form of the final results more closely resemble their 3D counterparts, in which the dielectric function is expressed in terms of the TO frequency. For small $q$,  the Born charges are (to lowest-order) constant, and so the LO phonon dispersion takes the form $\omega_{\mathbf{q},\mathrm{LO}}=\sqrt{\omega^2_{\mathrm{TO}}+2v_{\mathrm{g}}\omega_{\mathrm{TO}}q} \simeq \omega_{\mathrm{TO}} + v_{\mathrm{g}}q$, where $v_{\mathrm{g}}$, the LO phonon group velocity, is defined from microscopic parameters through the relation
 \begin{equation}
    v_{\mathrm{g}} = \frac{e^2 }{4\epsilon_0 \epsilon_{\mathrm{env}}\omega_{\Gamma, \mathrm{TO}}\Omega}\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \tilde{\mathbf{Z}}_\kappa\boldsymbol{\eta}_\kappa\Big|^2. 
\end{equation}
Thus, in the long wavelength limit, we have the following universal parameterization of the conductivity of a polar monolayer:
\begin{equation}
    \sigma(\omega) =  \frac{-4\iu\epsilon_0\epsilon_{\mathrm{env}}\omega\omega_{\mathrm{TO}}v_{\mathrm{g}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
    \label{eq:localsigma}
\end{equation}

@@

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $e^2\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \tilde{\mathbf{Z}}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 \equiv \frac{Q_*^2}{M_*}$, with $Q_*$ being an effective charge and $M_*$ being an effective mass.

Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$.

This squared frequency can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}"$  should be considered as 0. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c).

>>>>

% --- ABSTRACT ---
\begin{abstract}
    Extreme confinement of electromagnetic energy by phonon polaritons holds the promise of strong and new forms of control over the dynamics of matter. 
    To bring such control to the atomic-scale limit, it is important to consider phonon polaritons in two-dimensional (2D) systems. Recent studies have pointed out that in 2D, splitting between longitudinal and transverse optical (LO and TO) phonons is absent at the $\Gamma$ point, even for polar materials.  
    Does this lack of LO--TO splitting imply the absence of a phonon polariton in polar monolayers? 
    Here, we derive a first-principles expression for the conductivity of a polar monolayer specified by the wavevector-dependent LO and TO phonon dispersions. In the long-wavelength (local) limit, we find a universal form for the conductivity in terms of the LO phonon frequency at the $\Gamma$ point,
    its lifetime, and the group velocity of the LO phonon. Our analysis reveals that the phonon polariton of 2D is simply the LO phonon of the 2D system. For the specific example of hexagonal boron nitride (hBN), we estimate the confinement and propagation losses of the LO phonons, finding that high confinement and reasonable propagation quality factors coincide in regions which may be difficult to detect with current near-field optical microscopy techniques. Finally, we study the interaction of external emitters with two-dimensional hBN nanostructures, finding extreme enhancement of spontaneous emission due to coupling with localized 2D LO phonons, and the possibility of multi-mode strong and ultra-strong coupling between an external emitter and hBN phonons. This may lead to the design of new hybrid states of electrons and phonons based on strong coupling.

@@

% --- INTRODUCTION ---
Phonon-polaritons, hybrid quasiparticles of photons and optical phonons supported in polar materials, hold promise for nanoscale control of electromagnetic fields at mid-infrared and terahertz frequencies. Qualitatively, phonon polaritons share many features with plasmon polaritons in conductors. Recently, it has been shown that phonon polaritons enable confinement of light to volumes  $\sim10^6$ times smaller than that of a diffraction-limited photon in free-space~\cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra,li2018infrared,ma2018plane}. Due to this remarkable confinement and their relatively high lifetimes---around picoseconds---phonon polaritons open new opportunities for vibrational spectroscopy~\cite{autore2018boron}, radiative heat transfer~\cite{hillenbrand2002phonon}, and control of dynamics in quantum emitters~\cite{kumar2015tunable,rivera2017making,kurman2018control,flick2018strong}. 
Thus far, extreme confinement of phonon polaritons has been achieved by the use of thin-films (or nanostructuring), which shrink the in- and out-of-plane wavelength of polaritons with decreasing feature size (such as the film thickness)~\cite{dai2014tunable,dubrovkin2018ultra}. A monolayer is the ultimate limit of this effect. Concerning the optical response, the transition from three-dimensional (3D) to two-dimensional (2D) polar materials is nontrivial, however, since in a polar monolayer, the LO--TO splitting that gives rise to phonon polaritons in 3D is absent at the $\Gamma$ point~\cite{sanchez2002vibrational,mele2002electric,serrano2007vibrational,sohier2017breakdown}. This raises a fundamental question about the nature of  electromagnetic modes in polar monolayers.

@@

\section{Optical response of optical phonons in two-dimensions}
In this section, we develop a theory of electromagnetic response due to optical phonons in 2D systems. The key response function of interest is the conductivity of the monolayer. To that end, we consider the response of the ions of the monolayer due to an electric potential $\phi$. For that case, the interaction Hamiltonian is
\begin{equation}
    H_{\mathrm{int}} = \int \dd^2x ~\rho \phi = -\int \dd^2x~ (\nabla\cdot\mathbf{P})\phi,
    \label{eq:coupling}
\end{equation} 
with $\rho$ the induced charge density and $\mathbf{P}$ the induced polarization density associated with the ionic motion. Note that boldfaced quantities refer to vectors or tensors as appropriate. Within linear response theory, the polarization density can be straightforwardly evaluated from the displacement  $\mathbf{u}_{\kappa}$ of every atom $\kappa$ within the unit cell. Specifically, to first order, the polarization density is 
\begin{equation}
\mathbf{P} - \mathbf{P}^0 = \sum_\kappa(\mathbf{u}_\kappa\cdot\boldsymbol{\nabla}_{\mathbf{u}_\kappa})\mathbf{P} \equiv \frac{1}{\Omega}\sum_\kappa\mathbf{Z}_{\kappa}\mathbf{u}_{\kappa},
\end{equation} 
where $\mathbf{Z}_\kappa \equiv \Omega\boldsymbol{\nabla}_{\mathbf{u}_\kappa}\mathbf{P}$ is the Born effective charge tensor of ion $\kappa$ and $\Omega$ is the unit cell area. $\mathbf{P}^0$ is the equilibrium polarization in the absence of displacements, which is zero here. With this relation between polarization and ionic displacements, the interaction Hamiltonian in Equation~\eqref{eq:coupling} couples the scalar potential and the ionic displacements.  We consider the response of the monolayer to a potential of the form $\phi(\mathbf{r}) = \phi(\mathbf{q},\omega)\e^{\iu\mathbf{q}\cdot\mathbf{r}-\iu\omega t}$, where $\mathbf{q}$ is a 2D wavevector in the plane of the monolayer. Such a potential corresponds to a longitudinal electric field  $\mathbf{E}(\mathbf{r}) = \iu\mathbf{q}\phi(\mathbf{r})$.

In what follows, we assume the validity of the random-phase approximation (RPA) in calculating Coulombic interactions between ions in the polar lattice. Within the RPA, these Coulombic interactions are accounted for by taking the induced polarization $\mathbf{P}(\mathbf{q},\omega)$ to be proportional \emph{total} electric field, $\mathbf{E}_{\mathrm{tot}}(\mathbf{q},\omega)$, defined to be the sum of the externally applied electric field and the electric field created by the induced polarization. The polarization and total field are connected by the polarization-polarization response (tensor) function  $\boldsymbol{\Pi}(\mathbf{q},\omega)$ via
\begin{equation}
\mathbf{P}(\mathbf{q},\omega)  = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}_{\mathrm{tot}}(\mathbf{q},\omega),
\end{equation}
The polarization-polarization response function is related to the conductivity via the relation $\boldsymbol{\sigma}(\mathbf{q},\omega) = -\iu\omega\epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)$. From the Kubo formula, it follows that the conductivity is
\begin{equation}\label{eq:2dsusceptibility}
    \boldsymbol{\sigma}(\mathbf{q},\omega) =  \frac{-\iu \omega}{ \Omega\mathcal{Z}}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+\iu 0^+}\Big(\e^{-\beta E_m}-\e^{-\beta E_n} \Big),
\end{equation}
where $m,n$ are eigenstates in the phononic Fock space of the monolayer, $\mathbf{P}_{mn}(\mathbf{q}) \equiv \sum_\kappa \mathbf{Z}_\kappa\langle m|\mathbf{u}_\kappa(\mathbf{q})|n\rangle$ are matrix elements of the polarization associated with phonon modes (where $\mathbf{u}_\kappa(\mathbf{q})$ is the Fourier transform of the phonon displacement operator), $E_{m}$ ($E_n$) is the energy of state $m$ ($n$), $\beta \equiv 1/k_{\text{\textsc{B}}}T$ is the inverse temperature, and $\mathcal{Z}$ is the grand partition function. We now evaluate the contribution of optical phonons to the polarization-polarization response in the low-temperature limit $T \ll \hbar\omega_{\mathrm{ph}}/k_{\mathrm{B}}$ with $\omega_{\mathrm{ph}}$ a characteristic optical phonon frequency.  Considering the long-wavelength (small wavevector) limit, and taking a material with long-wavelength isotropy, such as hBN, we only have to consider the $qq$-component in the response tensor, where $qq$ denotes a pair of directions parallel to the wavevector. Denoting $\sigma_{qq}$ as simply $\sigma$, we find that the conductivity is given by \cite{rivera2018ab}: 
\begin{equation}
    \sigma(\mathbf{q},\omega) = \frac{-\iu\omega}{\hbar\Omega} \frac{2\omega_{\mathbf{q},\mathrm{L}}}{\omega^2_{\mathbf{q},\mathrm{L}}-\omega^2-\iu\omega\tau^{-1}}|\hat{\mathbf{q}}\cdot\langle 1_{\mathbf{q},\mathrm{L}}|\mathbf{P}(\mathbf{q})|0_{\mathbf{q},L}\rangle|^2,
    \label{eq:polpolresponse}
\end{equation}
where L-subscripts denote longitudinal polarization, $|0_{\mathbf{q},\mathrm{L}}\rangle$ ($|1_{\mathbf{q},\mathrm{L}}\rangle$) denotes a state with no (one) longitudinal phonon of wavevector $\mathbf{q}$, and $\hat{\mathbf{q}}$ denotes a unit vector in the direction of $\mathbf{q}$. We have also phenomenologically included the phonon dissipation rate $\tau^{-1}$, consistently with  a relaxation-time prescription.  The frequency $\omega_{\mathbf{q},\mathrm{L}}$ in the denominator, as in the case of bulk phonons, is the frequency of the longitudinal phonon of wavevector $\mathbf{q}$ prior to considering LO--TO splitting \cite{BornHuang:1954} (and near the $\Gamma$ point is approximately equal to the TO phonon frequency). This is consistent with the fact that LO--TO splitting is a collective effect arising from Coulomb interactions and the fact that the equation above represents a single-particle susceptibility. Coulomb interactions are accounted for in the random phase approximation, and to include them in the single-particle response amounts to an uncontrolled double-counting.

@@

While the conductivity is the main electromagnetic quantity of interest for electrodynamics applications, we briefly state the form of the (2D) permittivity, as its zeros immediately yield the longitudinal modes of the system, which are the LO phonons. The permittivity within the RPA, denoted $\epsilon_{\textrm{RPA}}$ is related to the polarization-polarization response function via \cite{jablan2009plasmonics} $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{1}{2}q\Pi(q,\omega)$ and the conductivity via $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{\iu q\sigma(q,\omega)}{2\omega\epsilon_0}$. Here, $\epsilon_{\mathrm{env}}$ is the average permittivity of the bulk above and below the monolayer, and is added to take into account the polarization arising from these bulk materials. Note that we have neglected any intrinsic high-frequency screening in the monolayer itself, which is only relevant for wavevectors comparable to the inverse layer spacing between monolayers. When considering non-local corrections to the conductivity at these large wavevectors, these must be taken into account \cite{sohier2016two,thygesen2017calculating,sohier2017breakdown}. Based on Equation~\eqref{eq:conductivity_simplified}, the zeros $\omega_{\mathbf{q}}$ of the RPA dielectric function satisfy:
\begin{equation}
    \omega^2_{\mathbf{q}} - \omega^2_{\mathrm{TO}} = \frac{V(q)}{e^2}\frac{1}{\Omega}q^2\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa}\Big|^2,
    \label{eq:zerosofresponse}
\end{equation}
  where $V(q)$ is the Coulomb interaction in Fourier space, which in two dimensions, is given by $V(q) = \frac{e^2}{2\epsilon_0\epsilon_{\mathrm{env}}q}$. Given that the zeros of the dielectric function are associated with longitudinal modes, one expects that $\omega_{\mathbf{q}}$ is in fact the frequency of the LO phonon mode. This is consistent with the result of Ref.~\citenum{sohier2017breakdown}, in which \relax{{}\ }\citeauthor{sohier2017breakdown} show that in 2D polar materials, the extra restoring forces on LO phonons relative to TO phonons, due to the Coulomb interaction, lead to a wavevector-dependent LO--TO splitting and zero LO--TO splitting at the $\Gamma$ point of the Brillouin zone.

Given these results, we now re-express the conductivity explicitly in terms of the 2D phonon dispersion, and derive a universal form for the conductivity in the local ($q\rightarrow 0$) limit specified in terms of three parameters: the LO phonon frequency at the $\Gamma$ point (i.e., $\omega_{\mathrm{TO}}$), the group velocity of the LO phonon at the $\Gamma$ point, and the damping rate. From Equation~\eqref{eq:zerosofresponse}, we can immediately write the conductivity as
  \begin{equation}
  \sigma(\mathbf{q},\omega) = \frac{-2\iu\epsilon_0\epsilon_{\mathrm{env}}\omega}{q}\frac{\omega^2_{\mathbf{q},\mathrm{LO}}-\omega^2_{\mathrm{TO}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
  \end{equation}
 In this expression, $\omega_\mathrm{L}$, the LO phonon frequency prior to LO-TO splitting, has been re-named as $\omega_{\mathrm{TO}}$ , the transverse optical phonon frequency, because in the absence of LO--TO splitting, they are degenerate. The RPA zeros $\omega_{\mathbf{q}$ have also been renamed as $\omega_{\mathbf{q},\mathrm{LO}}$. This is done in order to make the form of the final results more closely resemble their 3D counterparts, in which the dielectric function is expressed in terms of the TO frequency (see for example Eq. (\ref{eq:3deps})).

 For small $q$,  the Born charges are (to lowest-order) constant, and so the LO phonon dispersion takes the form $\omega_{\mathbf{q},\mathrm{LO}}=\sqrt{\omega^2_{\mathrm{TO}}+2v_{\mathrm{g}}\omega_{\mathrm{TO}}q} \simeq \omega_{\mathrm{TO}} + v_{\mathrm{g}}q$, where $v_{\mathrm{g}}$, the LO phonon group velocity, is defined from microscopic parameters through the relation
 \begin{equation}
    v_{\mathrm{g}} = \frac{\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 }{4\epsilon_0 \epsilon_{\mathrm{env}}\omega_{\mathrm{TO}}\Omega}. 
\end{equation}
Thus, in the long wavelength limit, we have the following universal parameterization of the conductivity of a polar monolayer:
\begin{equation}
    \sigma(\omega) =  \frac{-4\iu\epsilon_0\epsilon_{\mathrm{env}}\omega\omega_{\mathrm{TO}}v_{\mathrm{g}}}{\omega^2_{\mathrm{TO}}-\omega^2-\iu\omega\tau^{-1}}.
    \label{eq:localsigma}
\end{equation}

@@

The second comment is that much of what has been discussed here has a strong analogy with the theory of optical response in electron gases in 2D, and particularly the relation between plasmons in 2D and 3D. To elaborate on this analogy, we take Equation~\eqref{eq:zerosofresponse} in the case of a two-atom unit cell (such as hBN), and note that the term in the sum over Born charges can be written as $\Big|\hat{\mathbf{q}}\cdot \sum_\kappa \mathbf{Z}_\kappa\boldsymbol{\eta}_\kappa\Big|^2 \equiv \frac{Q_*^2}{M_*}$, with $Q_*$ being an effective charge and $M_*$ being an effective mass. Then, the LO--TO splitting can be written as $\omega^2_{\mathbf{q}} - \omega^2_{\mathbf{q},\mathrm{TO}} = \frac{Q_*^2}{2\epsilon_0\epsilon_{\mathrm{env}} M_*}q$. Now we note that the RHS is exactly the squared-frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ for a plasma oscillation in a 2D gas of charged particles with charge $Q_*$ and mass $M_*$. To connect to LO-TO splitting in phonons, this squared frequency $\omega^2_{\mathbf{q}\mathrm{p}}$ can be thought of as the ``LP--TP'' splitting between longitudinal and transverse plasma oscillations. Of course, there are no transverse plasma oscillations due to the structure-less nature of the electron, and so ``$\omega_{\mathrm{TP}}"$  should be considered as 0. In the three-dimensional plasmon case, ``$\omega_{\mathrm{TP}} = 0$'', but the difference between the squared longitudinal and transverse plasma oscillation frequencies at zero-wavevector is non-zero and given by $\omega_{\mathrm{p}}^2$. In other words, the plasma frequency in electron gases is analogous to the LO--TO splitting in polar materials. The change in the dispersion of 3D versus 2D plasmons, like 3D versus 2D LO phonons, arises from the change in dimensionality of the Coulomb interaction from 3D to 2D.  This analogy between phononic and plasmonic behavior as a function of dimension is illustrated schematically in Figures~\ref{fig:1}(b,c) in order to help unify the understanding of 2D plasmonics and phononics. These considerations should also extend to the one-dimensional case.
